---
author: "Arienne Calonge"
date: "2025-07-30"
title: "ICES NS-IBTS data availability"
output: 
  html_document:
    self_contained: true
runtime: shiny
---

## Harvesting data from ICES DATRAS
Setup libraries
```{r, echo=FALSE}
library(tidyverse)
library(icesDatras)
library(icesAdvice)
library(icesVocab)
library(mapplots)
library(leaflet)
library(shiny)
library(CoordinateCleaner)

setwd("C:/Users/arienne.calonge/OneDrive - VLIZ/Ari/DEMASK/R/datras")
```

## 1. Harvest data
```{r, echo=FALSE, eval=FALSE}
#find species codes in icesVocab

aphia_sp <- findAphia(c("Gadus morhua", "Melanogrammus aeglefinus", "Clupea harengus", "Dicentrarchus labrax", "Molva molva"), latin = TRUE)

names(aphia_sp) <- c("Gadus morhua", "Melanogrammus aeglefinus", "Clupea harengus", "Dicentrarchus labrax", "Molva molva")
aphia_lookup <- setNames(names(aphia_sp), aphia_sp)

#check availability of data from ICES DATRAS

    getDatrasDataOverview(surveys = 'NS-IBTS', long = TRUE)
    
    # Initialize empty list to store results
    all_catch <- list()
    
    # Loop through years
    for (yr in 2000:2024) {
      cat("Processing year:", yr, "\n")
      result <- getCatchWgt(
        survey = 'NS-IBTS',
        years = yr,
        quarters = c(1,3),
        aphia = c(126436, 126437, 126417, 126975, 126461)
      )
      all_catch[[as.character(yr)]] <- result
      print(yr)
    }
    
    # Merge all data frames into one
    CatchWgt <- do.call(rbind, all_catch)
    
    write_csv(CatchWgt, 'CatchWeight_200024.csv')

# Get length-based information such as measured length, individual counts, and sub-factors of sampled species

    HL = getHLdata(survey= 'NS-IBTS', year = 2023:2024, quarter = 1:4)

# adult and larvae?
```

## 2. Clean data 
```{r, echo=FALSE, eval=FALSE}
    colnames(CatchWgt)
    colSums(is.na(CatchWgt)) > 0
    
    # explore columns that are NA
    subset_NA <- CatchWgt[is.na(CatchWgt$HaulDUr), ]
    
    # retain only samples with haul durations of 25 to 35 mins and average haul depth of <= 250 m and eliminate duplicated occurrences
    range(CatchWgt$HaulDur)
    CatchWgt_proc = CatchWgt %>% filter(HaulDur >= 25 & HaulDur <=35) %>% filter(Depth <= 250) %>% distinct()
    CatchWgt_proc <- CatchWgt_proc[!is.na(CatchWgt_proc$CatchWgt), ]
    
    # remove if haul positions missing or same as shoot position
    CatchWgt_proc <- CatchWgt_proc[!(CatchWgt_proc$HaulLat == CatchWgt_proc$ShootLat & CatchWgt_proc$HaulLong == CatchWgt_proc$ShootLong), ]
    
    # check if haullong and haullat are in the correct ICES statrec
    CatchWgt_proc$CalculatedStatRec <- ices.rect2(CatchWgt_proc$HaulLong, CatchWgt_proc$HaulLat)
    wrong_statrec <- CatchWgt_proc[CatchWgt_proc$StatRec != CatchWgt_proc$CalculatedStatRec, ]
    
    CatchWgt_proc <- CatchWgt_proc[CatchWgt_proc$StatRec == CatchWgt_proc$CalculatedStatRec, ]
    
    CatchWgt_proc <- CatchWgt_proc[rowSums(is.na(CatchWgt_proc)) != ncol(CatchWgt_proc), ]
    
    CatchWgt_proc$SpeciesName <- aphia_lookup[as.character(CatchWgt_proc$Valid_Aphia)]
    
    #outlier check - coordinates
    CatchWgt_proc_ <- cc_outl(
                        CatchWgt_proc,
                        lon = "HaulLong",
                        lat = "HaulLat",
                        species = "SpeciesName",
                        method = "quantile",
                        mltpl = 1.5,
                        value = "flagged", #TRUE = test passed and FALSE = test failed/potentially problematic
                        sampling_thresh = 0,
                        verbose = TRUE,
                        min_occs = 7,
                        thinning = FALSE
                      )
    
    CatchWgt_proc_ <- ifelse(CatchWgt_proc_, "Passed", "Problematic")
    CatchWgt_proc$outlier_coord <- CatchWgt_proc_ 
    CatchWgt_proc <- CatchWgt_proc %>% filter(outlier_coord=="Passed")
    
    #outlier check - statistics
    
    #create unique Row ID
    CatchWgt_proc$RowID <- 1:nrow(CatchWgt_proc)
    
    # Function to detect outliers within each species group
    detect_outliers_iqr <- function(CatchWgt_proc) {
      Q1 <- quantile(CatchWgt_proc$CatchWgt, 0.25, na.rm = TRUE)
      Q3 <- quantile(CatchWgt_proc$CatchWgt, 0.75, na.rm = TRUE)
      IQR_val <- Q3 - Q1
      lower_bound <- Q1 - 1.5 * IQR_val
      upper_bound <- Q3 + 1.5 * IQR_val
      CatchWgt_proc %>% filter(CatchWgt < lower_bound | CatchWgt > upper_bound)
    }
    
    # Apply function to each species (grouped by Valid_Aphia)
    outliers_per_species <- CatchWgt_proc %>%
      group_by(Valid_Aphia) %>%
      group_modify(~ detect_outliers_iqr(.x))%>%
      select(RowID, Valid_Aphia, CatchWgt) %>%
      mutate(is_outlier = TRUE)

    # Flag outliers in the original dataframe
    CatchWgt_proc <- CatchWgt_proc %>%
        left_join(outliers_per_species %>% mutate(is_outlier = TRUE),
                  by = c("RowID", "Valid_Aphia", "CatchWgt")) %>%
        mutate(is_outlier = ifelse(is.na(is_outlier), FALSE, is_outlier))
    
    write_csv(CatchWgt_proc, 'CatchWeight_200024_proc.csv')

    save(CatchWgt_proc, file = "CatchWgt_proc.RData")
```

#Load cleaned data from 2015 - geographical outliers were removed, but statistical outliers (IQR) were only flagged in CatchWgt_proc.RData
```{r}
load("CatchWgt_proc.RData")

CatchWgt_proc = CatchWgt_proc %>% filter(Year >= 2015)
```

#Plot available data
```{r plot, echo=FALSE}
    # check completeness of data per species
    
    summary_data = CatchWgt_proc %>% group_by(Valid_Aphia, Year, Quarter) %>% summarize(NoHauls = n())
    
    quarter_coverage <- CatchWgt_proc %>%
      distinct(Valid_Aphia, Year, Quarter) %>%
      group_by(Valid_Aphia, Year) %>%
      summarize(QuartersPresent = n(), .groups = "drop")
    
    quarter_coverage <- quarter_coverage %>%
      mutate(Coverage = case_when(
        QuartersPresent == 2 ~ "Both",
        QuartersPresent == 1 ~ "One",
        TRUE ~ "None"
      ))
    
    # Get full species-year grid
    full_grid <- expand.grid(
      Valid_Aphia = unique(CatchWgt_proc$Valid_Aphia),
      Year = unique(CatchWgt_proc$Year)
    )
    
    # Join to add missing combinations as NA
    coverage_full <- full_grid %>%
      left_join(quarter_coverage, by = c("Valid_Aphia", "Year")) %>%
      mutate(Coverage = ifelse(is.na(Coverage), "None", Coverage))
    
    ggplot(coverage_full, aes(x = Year, y = as.factor(Valid_Aphia), fill = Coverage)) +
      geom_tile(color = "white") +
      scale_fill_manual(values = c("None" = "gray90", "One" = "orange", "Both" = "steelblue")) +
      labs(
        title = "Quarter 1 & 3 Coverage by Species and Year",
        x = "Year",
        y = "Species (Valid_Aphia)",
        fill = "Coverage"
      ) +
      theme_minimal() +
      theme(axis.text.y = element_text(size = 8))
    
```

```{r, echo=FALSE}
    
    # Plot catch weight per species over time
    
    CatchWgt_proc$Date <- as.Date(
      paste0(CatchWgt_proc$Year, "-", sprintf("%02d", CatchWgt_proc$Month),"-", CatchWgt_proc$Day),
      format = "%Y-%m-%d"
    )

    ggplot(CatchWgt_proc, aes(x = Date, y = CatchWgt, color = SpeciesName)) +
      geom_point(alpha = 0.6) +
      labs(title = "Catch Weight (kg)",
           x = "Date",
           y = "Catch Weight") +
      theme_minimal()
    
    # Plot with outliers
    ggplot(CatchWgt_proc, aes(x = Date, y = CatchWgt, color = SpeciesName)) +
      geom_point(aes(shape = is_outlier), alpha = 0.6, size = 2) +  # Shape by outlier label
      labs(title = "Catch Weight (kg)",
           x = "Time",
           y = "Catch Weight",
           shape = "Outlier Status") +
      theme_minimal()
          
```

## 3. Map occurrences 
Remove shoot/haul positions outside North Sea range -> not done! 

```{r, echo=FALSE}
message("Total no. of sampling points:",nrow(CatchWgt_proc))
message("No. of geographical outliers: ", 230) #sum(CatchWgt_proc_ == "Problematic", na.rm = TRUE)
message("No. of statistical outliers: ", sum(CatchWgt_proc$is_outlier == "TRUE", na.rm = TRUE))
```

```{r ui, echo=FALSE}
selectInput("SpeciesName", "Select species:", choices = unique(CatchWgt_proc$SpeciesName))

leafletOutput("map", height = 500)
```

```{r map, echo=FALSE}
output$map <- renderLeaflet({
  req(input$SpeciesName)

  filtered <- CatchWgt_proc %>%
    filter(SpeciesName == input$SpeciesName)

  leaflet(data = filtered) %>%
    addTiles() %>%
    addCircleMarkers(
      lng = ~HaulLong, lat = ~HaulLat,  # double-check if column is HaulLon or HaulLong
      popup = ~paste("Species:", SpeciesName),
      radius = 5,
      color = "blue"
    )
})
```

Static maps per species: Haddock
```{r, echo=FALSE}
# Get unique species names
species_list <- unique(CatchWgt_proc$SpeciesName)

leaflet(data = CatchWgt_proc %>% filter(SpeciesName==species_list[1])) %>%
  addTiles() %>%
  addCircleMarkers(
    lng = ~HaulLong,
    lat = ~HaulLat,
    radius = 2,
    color = "blue"
  )
```

Static maps per species: Cod
```{r, echo=FALSE}
# Get unique species names
species_list <- unique(CatchWgt_proc$SpeciesName)

leaflet(data = CatchWgt_proc %>% filter(SpeciesName==species_list[3])) %>%
  addTiles() %>%
  addCircleMarkers(
    lng = ~HaulLong,
    lat = ~HaulLat,
    radius = 2,
    color = "blue"
  )
```

Static maps per species: Ling
```{r, echo=FALSE}
# Get unique species names
species_list <- unique(CatchWgt_proc$SpeciesName)

leaflet(data = CatchWgt_proc %>% filter(SpeciesName==species_list[4])) %>%
  addTiles() %>%
  addCircleMarkers(
    lng = ~HaulLong,
    lat = ~HaulLat,
    radius = 2,
    color = "blue"
  )
```

Static maps per species: European seabass
```{r, echo=FALSE}
# Get unique species names
species_list <- unique(CatchWgt_proc$SpeciesName)

leaflet(data = CatchWgt_proc %>% filter(SpeciesName==species_list[5])) %>%
  addTiles() %>%
  addCircleMarkers(
    lng = ~HaulLong,
    lat = ~HaulLat,
    radius = 2,
    color = "blue"
  )
```

Map statistical outliers based on IQR
```{r, echo=FALSE}
outliers_map_data <- CatchWgt_proc %>%
  filter(is_outlier == TRUE, !is.na(HaulLat), !is.na(HaulLong))

# Create the leaflet map
leaflet(data = outliers_map_data) %>%
  addTiles() %>%
  addCircleMarkers(
    lng = ~HaulLong,
    lat = ~HaulLat,
    radius = 5,
    color = "red",
    stroke = FALSE,
    fillOpacity = 0.8,
    popup = ~paste("Species:", SpeciesName, "<br>",
                   "Catch Weight:", CatchWgt, "kg")
  )
```
