---
author: "Jo-Hannes Nowé"
date: "2025-08-28"
title: "ICES datras analysis"
format:
  html:
    self-contained: true
    embed-resources: true
    fig-format: png
    code-fold: false
---

## Harvesting data from ICES DATRAS

Setup libraries

```{r, echo=FALSE}
library(sf)
library(ggplot2)
library(terra)
library(tidyverse)
library(cmocean)
```
```{r}
#Setup directories
ifelse(!dir.exists("figures"), dir.create("figures"), FALSE)
```


## 1. Load data

```{r}
#CatchWgt_proc
load("CatchWgt_proc.RData")
head(CatchWgt_proc)
#First try out for Cod
species <- "Gadus morhua"
CatchWgt_proc <- CatchWgt_proc %>%
  dplyr::filter(SpeciesName == species)
```

For the moment we will work with the Catchweight. Arienne is now turning that into Catch per Unit Effort per swept area.

## Making the grid cells

### Download the study area (North Sea)

See code chunk we had to download the ospar data.

```{r}
load_ospar <- function(regions=c("I","II","III","IV","V"), filepath){
  if(!file.exists(filepath)){
    url <- "https://odims.ospar.org/public/submissions/ospar_regions/regions/2017-01/002/ospar_regions_2017_01_002-gis.zip"
    download.file(url,file.path(dirname(filepath),"ospar_REGIONS.zip"),mode="wb")
    unzip(zipfile=file.path(dirname(filepath),"ospar_REGIONS.zip"),exdir=dirname(filepath))
  }
  ospar_regions <- st_read(filepath)
  #Subsetting to regions of interest
  study_area <- ospar_regions[ospar_regions$Region %in% regions,]
  study_area <- sf::st_make_valid(study_area)%>%st_union()
  return(study_area)
}
```

```{r}
north_sea <- load_ospar(regions = "II", #OSPAR region II is the Greater North Sea
           filepath = "study_area/ospar_regions_2017_01_002.shp")
ggplot(north_sea)+
  geom_sf()
```

### Turn the study area into a raster with grid cells

We first run the methodology for a coarser grid to more clearly show the different
steps. Later we perform the analysis for the 10x10km grid.

```{r}
#Turn the shapefile into spatVector to be able to use the terra package
north_sea_vect <- terra::vect(north_sea)
terra::plot(north_sea_vect)

#Turn the spatVector into a raster
r <- terra::rast(north_sea_vect,
                 resolution = 0.12, #Corresponds roughly to 10x10km
                 vals = 1)%>%
  terra::mask(north_sea_vect)

#Mockup example with coarser resolution to show methodology
r <- terra::rast(north_sea_vect,
                 resolution = 1, #Corresponds roughly to 10x10km
                 vals = 1)%>%
  terra::mask(north_sea_vect)

```

### Plot the rasters

```{r}
outlines <- mean(r) %>% as.polygons(dissolve = FALSE)

png(file.path("figures","grid_coarse.png"), width = 1200, height = 1800, res = 300)
terra::plot(north_sea_vect)
terra::plot(outlines, add = TRUE)
dev.off()
```

See code chunk we had from plotting the raster with the grid cell lines.

## Keep grid cells with data in at least \>90% of the years

### Turn the dataframe into spatial points

The spatial points should have geometry, speciesName, Catchwgt, year, month

```{r}
catch_all <- CatchWgt_proc %>%
  dplyr::select(Latitude = ShootLat, #Take this or HaulLat? Or mean of these?
                Longitude = ShootLong,
                Year, Month, SpeciesName, CatchWgt)%>%
  dplyr::mutate(quarter = lubridate::quarter(Month))

#First try it for one quarter and year
q = 1
y = 2000
catch <- catch_all%>%
  dplyr::filter(quarter == q,
                Year == y)
catch_sf <- st_as_sf(catch, coords = c("Longitude", "Latitude"), crs = "EPSG:4326")

png(file.path("figures","sampling_coarse.png"), width = 1200, height = 1800, res = 300)
ggplot(catch_sf)+
  geom_sf(data = north_sea)+
  geom_sf()+
  labs(title = "Catch weight sampling points",
       subtitle = paste0(species," Year: ",y,", Q",q))
dev.off()
```

### Turn the spatial points into yearly rasters

We will do this first manually for a coarser resolution and one quarter-year combination to show the methodology with plots.

```{r}
# Setup color scheme using the cmocean package
algae <- cmocean("algae")
pal <- algae(15)
```

```{r}
#Get the coordinates of the center of each grid cell
cells_not_na <- cells(r)
center_grid <- xyFromCell(r, cells_not_na)

#Create a SpatVector from these center points
center_grid <- terra::vect(center_grid, crs = "EPSG:4326")

#Plot the center points in the study area
png(file.path("figures","center_grid_coarse.png"), width = 1200, height = 1800, res = 300)
ggplot(data = st_as_sf(center_grid))+
  geom_sf(data = north_sea) +
  geom_sf()+
  labs(title = "Center of grid cells")
dev.off()

#Transform to projected crs so we can take the buffer in km instead of degrees
center_grid_proj <- terra::project(center_grid, "EPSG:25832")

#Draw a buffer around the centers of 65km
center_buffer <- terra::buffer(center_grid_proj, width = 65000)
#Project back to same CRS as catchweight points
center_buffer <- terra::project(center_buffer, "EPSG:4326")
#Turn into sf object for easier spatial operations
center_buffer <- st_as_sf(center_buffer)

#Plot the buffers in the study area
png(file.path("figures","buffer_grid_coarse.png"), width = 1200, height = 1800, res = 300)
ggplot(center_buffer)+
  geom_sf(data = north_sea)+
  geom_sf(alpha = 0.1)+
  geom_sf(data = st_as_sf(center_grid))+
  labs(title = "Buffer of grid cells")
dev.off()

#Plot the sampling points in the buffers
png(file.path("figures","buffer_points_coarse.png"), width = 1800, height = 1800, res = 300)
ggplot(center_buffer)+
  geom_sf(data = north_sea)+
  geom_sf(alpha = 0.1)+
  geom_sf(data = st_as_sf(center_grid), aes(colour = "Grid center"))+
  geom_sf(data = catch_sf, aes(colour = "Sampling point")) +
  scale_color_manual(
    values = c("Grid center" = "black", "Sampling point" = "red"),
    name = "Points")+
  labs(title = "Buffer with sampling points",
       subtitle = paste0(species," Year: ",y,", Q",q))
dev.off()

#Check which points intersect with which polygons of buffers
contains_list <- st_contains(center_buffer, catch_sf)
#Calculate the value of each grid cell as the mean of values around the center
mean_catch <- sapply(contains_list, function(idx) {
  if (length(idx) == 0) {
    return(NA)  # no points in this buffer
  } else {
    mean(catch_sf$CatchWgt[idx], na.rm = TRUE)
  }
})

#Each buffer now contains the mean catch value
center_buffer$mean_catch <- mean_catch
#Plot the value of each buffer
png(file.path("figures","buffer_values_coarse.png"), width = 1800, height = 1800, res = 300)
ggplot(data = center_buffer, alpha = 0.01)+
  geom_sf(aes(fill = mean_catch))+
  geom_sf(data = north_sea, fill = NA)+
  scale_fill_gradientn(colours = pal, na.value = "transparent")+
  geom_sf(data = st_as_sf(center_grid))+
  labs(title = "Buffer with calculated mean catch",
       subtitle = paste0(species," Year: ",y,", Q",q))
dev.off()

#Give each center of the grid cells this value for easier rasterizing
center_grid$mean_catch <- mean_catch

#Also plot the center of each of the grid cells with their value
png(file.path("figures","center_values_coarse.png"), width = 1800, height = 1800, res = 300)
ggplot(data = st_as_sf(center_grid), alpha = 0.01)+
  geom_sf(aes(colour = mean_catch))+
  geom_sf(data = north_sea, fill = NA)+
  scale_color_gradientn(colours = pal, na.value = "transparent")+
  labs(title = "Center grid cell with calculated mean catch",
       subtitle = paste0(species," Year: ",y,", Q",q))
dev.off()

#Rasterize mean catchweight
r_mean <- rasterize(x = center_grid,
                    y = r, #grid of pre-defined resolution within Greater North Sea
                    field = "mean_catch",
                    fun = "first")

#Plot the raster
r_mean_df <- as.data.frame(r_mean, xy = TRUE) %>%
  rename(mean_catch = 3)
png(file.path("figures","raster_coarse.png"), width = 1800, height = 1800, res = 300)
ggplot() +
  geom_raster(data = r_mean_df, aes(x = x, y = y, fill = mean_catch))+
  geom_sf(data = north_sea, fill = NA)+
  scale_fill_gradientn(colours = pal, na.value = "transparent")+
  labs(title = "Raster with calculated mean catch",
       subtitle = paste0(species," Year: ",y,", Q",q),
       x = NULL, y = NULL)
dev.off()
```

We make a function to loop over every quarter-year-species combination.

```{r raster_function}
calculate_mean <- function(r, catch_sf){
#Get the coordinates of the center of each grid cell
cells_not_na <- cells(r)
center_grid <- xyFromCell(r, cells_not_na)

#Create a SpatVector from these center points
center_grid <- terra::vect(center_grid, crs = "EPSG:4326")

#Transform to projected crs so we can take the buffer in km instead of degrees
center_grid_proj <- terra::project(center_grid, "EPSG:25832")

#Draw a buffer around the centers of 65km
center_buffer <- terra::buffer(center_grid_proj, width = 65000)
#Project back to same CRS as catchweight points
center_buffer <- terra::project(center_buffer, "EPSG:4326")
#Turn into sf object for easier spatial operations
center_buffer <- st_as_sf(center_buffer)

#Check which points intersect with which polygons of buffers
contains_list <- st_contains(center_buffer, catch_sf)
#Calculate the value of each grid cell as the mean of values around the center
mean_catch <- sapply(contains_list, function(idx) {
  if (length(idx) == 0) {
    return(NA)  # no points in this buffer
  } else {
    mean(catch_sf$CatchWgt[idx], na.rm = TRUE)
  }
})

#Each buffer now contains the mean catch value
center_buffer$mean_catch <- mean_catch

#Give each center of the grid cells this value for easier rasterizing
center_grid$mean_catch <- mean_catch

#Rasterize mean catchweight
r_mean <- rasterize(x = center_grid,
                    y = r, #grid of pre-defined resolution within Greater North Sea
                    field = "mean_catch",
                    fun = "first")
return(r_mean)
}
```

We make a spatraster where each layer is a different quarter-year combination.

```{r}
#We choose a finer raster now to fit the project needs of 10x10km
r <- terra::rast(north_sea_vect,
                 resolution = 0.12, #Corresponds roughly to 10x10km
                 vals = 1)%>%
  terra::mask(north_sea_vect)
```

```{r}
#Define all quarter-year combinations
q_yr_combo <- expand.grid(quarter = unique(catch_all$quarter),
            Year = unique(catch_all$Year)) %>%
  arrange(Year, quarter)
```


```{r}
#Run the analysis for each quarter-Year combination
mean_raster <- apply(q_yr_combo,
                     1, #to apply it row by row 
                     function(x) {
                       q <- x["quarter"]
                       y <- x["Year"]
                      catch_sf <- catch_all%>%
  dplyr::filter(quarter == q,
                Year == y)%>%
                        st_as_sf(coords = c("Longitude", "Latitude"), crs = "EPSG:4326")
calculate_mean(r, catch_sf) 
                     })
#Turn the list of rasters into a spatraster with different layers
mean_raster <- terra::rast(mean_raster)
terra::writeCDF(mean_raster, paste0("mean_raster_", 
                                    tolower(gsub(" ", "",species)),".nc"), overwrite=TRUE) #remove the space from species name
```

Split the spatrasters per quarter to check the different years for each quarter.
```{r}
n_years <- length(unique(catch_all$Year))

#We split every Q1 from Q3
split_list <- rep(1:2, times = n_years)
raster_split <- terra::split(mean_raster, split_list)
raster_winter <- raster_split[[1]]
names(raster_winter) <- unique(catch_all$Year)
raster_summer <- raster_split[[2]]
names(raster_summer) <- unique(catch_all$Year)
```

Making plots

```{r}
plot_mean_raster <- function(mean_raster, quarter, species){
#Plot the raster
upper_limit <- 200000
mean_raster_df <- as.data.frame(terra::clamp(mean_raster, upper = upper_limit, values=TRUE), #put an upper limit on values for plotting purposes
                                  xy = TRUE) %>%
  tidyr::pivot_longer(cols = "2015":"2024",
                      names_to = "Year",
                      values_to = "mean_catch")
png(file.path("figures",paste0("raster_",
tolower(gsub(" ", "",species)),"Q", as.character(quarter) ,".png")), width = 3200, height = 1800, res = 500)
p <- ggplot() +
  geom_raster(data = mean_raster_df, aes(x = x, y = y, fill = mean_catch))+
  geom_sf(data = v, fill = NA)+
  scale_fill_gradientn(colours = pal, na.value = "transparent",
                       limits = c(0, 200000),
                       breaks = seq(0,200000, by = 40000))+
  labs(title = "Raster with calculated mean catch",
       subtitle = paste0(species, " Q",as.character(quarter)),
       x = NULL, y = NULL)+
  facet_wrap("Year",ncol = 5)
print(p)
dev.off()
}
```

```{r}
plot_mean_raster(mean_raster = raster_winter,
                 quarter = 1,
                 species = species)
plot_mean_raster(mean_raster = raster_summer,
                 quarter = 3,
                 species = species)
```


### Turn the cells with lower coverage then 0.90 of total years to NA

First we only take the years 2015-2024
```{r}
raster_winter <- raster_winter[[16:nlyr(raster_winter)]]

raster_summer
```
We put the values of grid cells that have NA values in a certain amount of years to NA.
This ensures we only look at areas that are well studied over the different years.


```{r}
# First we define this threshold as a percentage.
thr_year <- 0.90 * nlyr(raster_winter) #Percentage of years that should have data

# count how many layers are non-NA in each cell
valid_counts <- app(raster_winter, function(x) sum(!is.na(x)))
terra::plot(valid_counts, col = pal)
terra::plot(valid_counts >=thr_year)
# mask out cells where count < threshold
r_masked <- mask(raster_winter, valid_counts >= thr_year, maskvalues=FALSE,updatevalue = NA)
names(r_masked) <- seq(from = 2015, to= 2024, by = 1)
terra::plot(r_masked)
mean_raster_df <- as.data.frame(terra::clamp(r_masked, upper = 200000, values=TRUE), #put an upper limit on values for plotting purposes
                                  xy = TRUE) %>%
  tidyr::pivot_longer(cols = "2015":"2024",
                      names_to = "Year",
                      values_to = "mean_catch")
# png(file.path("figures",paste0("raster_",
# tolower(gsub(" ", "",species)),"Q", as.character(quarter) ,".png")), width = 3200, height = 1800, res = 500)
p <- ggplot() +
  geom_raster(data = mean_raster_df, aes(x = x, y = y, fill = mean_catch))+
  geom_sf(data = north_sea, fill = NA)+
  scale_fill_gradientn(colours = pal, na.value = "transparent",
                       limits = c(0, 200000),
                       breaks = seq(0,200000, by = 40000))+
  labs(title = "Raster with calculated mean catch",
       subtitle = "Only grid cells with data in 90% of the years",
       x = NULL, y = NULL)+
  facet_wrap("Year",ncol = 5)
print(p)
```

Next step is to calculate the grid cells needed to make up 50% of the values.

```{r}
terra::plot(r_masked[[1]], col= pal)
#Choose the threshold for the cumulative percentage
thr_percent <- 0.5
#First we order the values from large to small
#Extract the values
vals <- as.data.frame(r_masked, cells = TRUE, na.rm = FALSE)

for(i in 2:ncol(vals))
  #Index the cell column and one of the years
  df_yr <- vals[,c(1,i)]
  #Order based on the year value
  df_ordered <- df_yr[order(df_yr[,2], decreasing = TRUE),]
  # Calculate cumsum
  df_ordered$c_sum <- cumsum(df_ordered[,2])
  total <- sum(df_ordered[,2], na.rm = TRUE)
  df_ordered$c_percent <- df_ordered$c_sum / total
  # Flag whenever we reach 50
  df_ordered$top <- df_ordered$c_percent <= thr_percent
  #Check how to include cell that crosses 50%
  
  #Create a mask based on these true false
  mask_raster <- rast(r_masked[[i-1]])
  values(mask_raster) <- NA
  values(mask_raster)[df_ordered$cell[df_ordered$top]] <- 1
  terra::plot(mask_raster)
  
  
  terra::plot(terra::vect(north_sea%>%st_transform("EPSG:3035")))
  terra::plot(terra::mask(r_masked[[8]], mask_raster), col = pal, add = TRUE)
  

```

```{r}
#install.packages('leastcostpath')
library(leastcostpath)
```



