---
title: "Test_threshold"
author: "Jo-Hannes Nowé"
format: html
editor: visual
---

This script is to show how we make plots when we have the area that exceeds the chosen threshold value.

```{r}
library(sf)
library(here)
library(terra)
library(ggplot2)
```

```{r}
study_area <- st_read(here("data",
                                "study_area",
                                "20250912_10x10km_grid.shp"))
r <- rast(x = terra::vect(study_area), resolution = 10000)
v <- rasterize(study_area, r, field = runif(14400))
plot(v)
```

Make the polygon that shows the area where the threshold was exceeded, can also be a raster first that we then turn into a vector layer.

```{r}
# create bbox
bbox <- st_bbox(
  c(
    xmin = 3750000,
    xmax = 4000000,
    ymin = 3400000,
    ymax = 3600000
  ),
  crs = st_crs(3035)
)

# convert bbox to polygon sf object
rectangle <- st_as_sfc(bbox) |>
  st_sf()

# inspect
ggplot(study_area)+
  geom_sf()+
  geom_sf(data = rectangle)
```

The rectangle is the area where the threshold was exceeded.

Run the analysis_fish_interpolation.qmd for 'Gadus morhua' until you get to the last part so that you have the core area_q1 raster

```{r}
shp <- "gadus_morhua_spawning_imr.shp"
scoring_raster(core_area_q1, species = params$species,
               quarter = "1",
               shp= paste0(stringr::str_to_lower(params$species)%>%
  stringr::str_replace(" ","_"),"_spawning_imr.shp"))
  consistent_core <- app(core_area_q1, function(x) sum(!is.na(x)))
  # Make zero-count cells transparent
  consistent_core[consistent_core == 0] <- NA
  consistent_core[consistent_core == 1] <- 1
  consistent_core[consistent_core %in% 2:3] <- 2
  consistent_core[consistent_core %in% 4:7] <- 3
  consistent_core[consistent_core %in% 8:10] <- 4
  
  if(is.null(shp) || !file.exists(here("data", "spawning", shp))){
    n_classes <-4
  print("No valid spawning shapefile was found.")
  } else{
    #Turn the spawning shapefile into a raster
    spawning <- st_read(here("data", "spawning", shp), quiet = TRUE)
    spawning <- st_transform(spawning, terra::crs(consistent_core))
    spawning$spawning <- TRUE
    spawning_raster <- terra::rasterize(terra::vect(spawning), consistent_core, field = "spawning", background = NA, touches = TRUE)
    names(spawning_raster) <- "spawning"
    consistent_core <- terra::mask(consistent_core,
                                   mask = spawning_raster,
                                   maskvalues = TRUE,
                                   updatevalue = 5)
    n_classes <- 5
  }
terra::plot(consistent_core)

```

Now we overlay the rectangle of the area with our consistent core scoring:

I'll try to plot the raster a bit less clear and then within the rectangle highlight it by giving opacity = 1

```{r}
#ORIGINAL RASTER into df
raster_df <- as.data.frame(consistent_core,xy=TRUE)%>%
  dplyr::rename(score = lyr.1)%>%
  dplyr::mutate(score = factor(score))

#Mask the raster to only this spot
masked_core <- terra::mask(consistent_core, rectangle)
#turn into df
masked_df <- as.data.frame(masked_core, xy = TRUE)%>%
  dplyr::rename(score = lyr.1)%>%
  dplyr::mutate(score = factor(score))

ggplot(data = study_area) +
    geom_raster(data = raster_df, aes(x = x, y = y, fill = score), alpha = 0.4)+
  geom_sf(data = rectangle, fill = NA)+
  geom_raster(data = masked_df, aes(x = x, y = y, fill = score), alpha = 1)
  
```

```{r}
ggplot(data = masked_df, aes(score))+
  stat_count()
```
